---
title: "Lexis diagrams"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_document

vignette: >
  %\VignetteIndexEntry{Lexis diagrams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`LexisPlot()` creates a blank parallelogram-shaped plot, based on the Lexis
diagram used in demography for age–period–cohort (APC) analysis.
Data can be added with `LexisPoints()`, `LexisLines()`, `LexisText()`, and
`LexisPolygon()`; the generic dispatcher `AddToLexis()` accepts any base
graphics plotting function.

A standard ternary triangle constrains three axes to the same range
(a + b + c = 1).  A Lexis plot relaxes this constraint: two of the three axes
are "bounded" and the third is derived automatically.  Specify ranges for any
two axes — `aRange`, `bRange`, `cRange` — and the third will be computed.

# A basic Lexis plot

With the default ranges (b ∈ [0, 1], c ∈ [0, 1], a derived) the plot
is a rhombus — equivalent to two ternary triangles sharing an edge:

```{r basic, fig.asp = 1}
library("Ternary")
par(mar = rep(0.2, 4))

LexisPlot()
```

# Age–Period–Cohort layout

In demographic applications, **Age + Cohort = Period**.  To produce the
conventional Lexis surface with horizontal Age lines and slanted Period edges,
specify `aRange` (Age) and `bRange` (Period), leaving Cohort as the derived
axis:

```{r apc, fig.asp = 0.7}
par(mar = rep(0.2, 4))

LexisPlot(
  aRange = c(0, 100),
  bRange = c(1900, 2000),
  alab = "Age", blab = "Period", clab = "Cohort",
  free.fun = function(a, b) b - a,
  grid.lines = 10, grid.minor.lines = 0,
  col = "white"
)

# Add some data points
set.seed(4781)
n <- 30
ages <- runif(n, 5, 90)
periods <- runif(n, 1910, 1990)
LexisPoints(a = ages, b = periods, pch = 16, col = "#22558866", cex = 1.2)
```

Three families of grid lines appear:
Age lines run horizontally, Period lines slant at 60° to the upper right,
and the third family slants at 60° to the upper left.

The `free.fun` argument controls how the derived axis is labelled on the
remaining two edges.
Passing `free.fun = function(a, b) b - a` labels them with
Period − Age, i.e. the demographic Cohort.
Without `free.fun`, labels show the ternary complement
(the value of the internal third coordinate).

# Asymmetric ranges (a derived)

When `bRange` and `cRange` are specified, `a` is the derived axis.
Asymmetric ranges produce a parallelogram wider than it is tall (or vice
versa):

```{r asymmetric, fig.asp = 0.6}
par(mar = rep(0.2, 4))

LexisPlot(
  bRange = c(1900, 2100),
  cRange = c(0, 99),
  blab = "Period", clab = "Age", alab = "Cohort",
  atip = "Cohort", btip = "Period", ctip = "Age",
  grid.lines = 10
)
```


# Adding data

Points are specified using the two bounded axes by name.  The convenience
wrappers `LexisPoints()`, `LexisLines()`, etc. handle coordinate
conversion automatically:

```{r data-points, fig.asp = 0.6}
par(mar = rep(0.2, 4))

LexisPlot(
  bRange = c(1900, 2100),
  cRange = c(0, 99),
  blab = "Period", clab = "Age", alab = "Cohort",
  grid.lines = 10
)

set.seed(4781)
n <- 40
period <- runif(n, 1920, 2080)
age <- runif(n, 5, 90)

LexisPoints(b = period, c = age, pch = 16, col = "#22558866", cex = 1.2)
```

# Compatibility with standard ternary functions

Because the underlying coordinate maths is the same, `TernaryPoints()`,
`TernaryLines()`, and other standard ternary functions also work on a
Lexis plot — simply pass all three coordinates (a, b, c):

```{r ternary-compat, fig.asp = 1}
par(mar = rep(0.2, 4))

LexisPlot(blab = "b", clab = "c", alab = "a")

# Both calls place a point at the same position:
LexisPoints(b = 0.3, c = 0.5, pch = 16, cex = 2, col = "red")
TernaryPoints(c(1 - 0.3 - 0.5, 0.3, 0.5), pch = 1, cex = 3, col = "blue")
```

The red dot (`LexisPoints`) and the blue circle (`TernaryPoints`) overlap
exactly.

# Orientation

`LexisPlot()` supports four orientations via the `point` parameter:

```{r orientation, fig.asp = 1, fig.width = 7}
par(mfrow = c(2, 2), mar = rep(1, 4))
for (dir in c("up", "right", "down", "left")) {
  LexisPlot(point = dir)
  text(mean(par("usr")[1:2]), mean(par("usr")[3:4]),
       paste0('point = "', dir, '"'), cex = 1.5)
}
```


# Colour fill and contours

`LexisPointValues()` evaluates a function at triangle tile centres across the
parallelogram, and `ColourLexis()` paints those tiles.  The function's
argument names must match the two bounded axes.

Use the `panel.first` argument to place colour behind the grid lines.
Because `panel.first` is evaluated inside `LexisPlot()`, call
`LexisPointValues()` from a plot with the same ranges first:

```{r colour-contour, fig.asp = 0.6}
par(mar = rep(0.2, 4))

# Set up ranges so LexisPointValues() can read them
MySurface <- function(b, c) sin(b / 30) * cos(c / 15)

LexisPlot(
  bRange = c(1900, 2100), cRange = c(0, 100),
  blab = "Period", clab = "Age", alab = "Cohort",
  grid.lines = 10, grid.minor.lines = 0
)

vals <- LexisPointValues(MySurface, resolution = 24)
ColourLexis(vals, legend = TRUE, bty = "n")
LexisContour(MySurface, resolution = 48)
```


# Where next

For other examples of usage, see the accompanying vignettes:

- [Introduction for new R users](new-users.html)

- [Getting started guide](Ternary.html)

- [Interpolating and contouring](interpolation.html)

- [Plotting Holdridge climate zones](Holdridge.html)
