---
title: "Creating ternary plots with Ternary"
author: "Martin R. Smith"
date: "23 November 2017"
output: rmarkdown::html_document

vignette: >
  %\VignetteIndexEntry{Creating ternary plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

I hope that this package proves simple to use.

Install the package with:

```{r Load package, eval=FALSE}
install.packages('Ternary')
```

Or if you want the latest development version of the package:

```{R Github package, eval=FALSE}
if (!require('devtools')) install.packages('devtools')
install_github('ms609/Ternary')
```

Once the package is installed, load it into the current R session with 
```{R Load}
library('Ternary')
```

There are two stages to creating a ternary plot: first, rendering the plot, styled
as you like it and pointing in any of the four compass directions; secondly, adding data.


## Create a blank plot
```{R Create blank plot, fig.width=7, fig.height=7}
par(mfrow=c(2, 2), mar=rep(0.5, 4))
for (dir in c('up', 'right', 'down', 'le')) {
  TernaryPlot(point=dir, 'A', 'B', 'C')
  TernaryText(list(A=c(10, 01, 01), B=c(01, 10, 01), C=c(01, 01, 10)), col=cbPalette8[4], font=2)
}
```

## Plot two stylised plots side by side, and plot data
```{R Do plotting}
par(mfrow=c(1, 2), mar=rep(0.3, 4))
TernaryPlot(' Redder', ' Greener', 'Bluer', point='right', lab.cex=0.8,
            grid.lty='solid', col=rgb(0.9, 0.9, 0.9), grid.col='white', 
            axis.col=rgb(0.6, 0.6, 0.6), ticks.col=rgb(0.6, 0.6, 0.6),
            padding=0.08)
data_points <- list(
  R = c(255, 0, 0), 
  O = c(240, 180, 52),
  Y = c(210, 222, 102),
  G = c(111, 222, 16),
  B = c(25, 160, 243),
  I = c(92, 12, 243),
  V = c(225, 24, 208)
)
AddToTernary(points, data_points, bg=vapply(data_points, function (x) rgb(x[1], x[2], x[3], 128, maxColorValue=255), character(1)), pch=21, cex=2.8)
AddToTernary(text, data_points, names(data_points), cex=0.8, font=2)
legend('bottomright', 
       pch=21, pt.cex=1.8,
       pt.bg=c(rgb(255, 0, 0,   128, NULL, 255), 
             rgb(240, 180,  52, 128, NULL, 255),
             rgb(210, 222, 102, 128, NULL, 255),
             rgb(111, 222,  16, 128, NULL, 255)),
       legend=c('Red', 'Orange', 'Yellow', 'Green'), 
       cex=0.8, bty='n')

TernaryPlot('Steam', 'Ice', 'Water', grid.lines=5)
#HorizontalGrid()
middle_triangle <- matrix(c(
  30, 40, 30,
  30, 30, 40,
  55, 20, 25
), ncol=3, byrow=TRUE)
TernaryPolygon(middle_triangle, col='#aaddfa', border='grey')
TernaryLines(list(c(0, 100, 0), middle_triangle[1, ]), col='grey')
TernaryLines(list(c(0, 0, 100), middle_triangle[2, ]), col='grey')
TernaryLines(list(c(100, 0, 0), middle_triangle[3, ]), col='grey')
```


It is also possible to use cartesian coordinates to plot onto the graph.

By default, the plotting area is a 1x1 square.

```{r Cartesian}
TernaryPlot(point='right')
cat("X range in this orientation:", TernaryXRange())
cat("\nY range in this orientation:", TernaryYRange())
arrows(x0=0.5, y0=0.4, x1=sqrt(3)/2, y1=0.4, length=0.1, col=cbPalette8[2])
text(x=mean(c(0.5, sqrt(3)/2)), y=0.4, "Increasing X", pos=3, col=cbPalette8[2])
text(x=0.5, y=0, "(0.5, 0)", col=cbPalette8[3])
text(x=0.8, y=-0.5, "(0.8, -0.5)", col=cbPalette8[3])
```


## Bugs and feature requests

I hope that the package comes in useful.  If there's anything it can't do that you wish
it could, please let me know by opening a [Github issue](https://github.com/ms609/Ternary/issues).

